\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-01-20 jasa/paper-jasa.tex
formatting of the tex file

@@

@@

@@

\begin{proposition} \label{thm:best-classifier}
Let $\mathbf{X} = (X_1, ..., X_p)$ and $Y$ be random input and output variables
with values in ${\cal X} \subseteq \mathbb{R}^p$\footnote{Without loss of
generality, the discussion assumes continuous random variables, but can be
extended to random variables of mixed types with the Dirac delta function.}
and ${\cal Y} = \{0, 1\}$ and mixed joint probability density  function
$p_{\mathbf{X},Y}(\mathbf{x}, y)$\footnote{Defined as
$p_{\mathbf{X},Y}(\mathbf{x}, y) = p_{\mathbf{X}|Y}(\mathbf{x}|y) P(Y=y) =
P(Y=y|X=x) p_{\mathbf{X}}(\mathbf{x})$.}. For the squared error loss, the best
regression function $s : {\cal X} \mapsto [0, 1]$, or equivalently the best
probabilistic classifier, is
\begin{equation}
s^*(\mathbf{x}) = \frac{P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{P(Y=0) p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
\end{equation}
\end{proposition}

@@

Equivalently, since $m^{-1}(s^*(\mathbf{x})) = \frac{1 - s^*(\mathbf{x})}{s^*(\mathbf{x})} = r(\mathbf{x})$,
Proposition~\ref{thm:best-classifier} shows that the density ratio estimation problem is strictly
equivalent to probabilistic classification, and that algorithms proposed to
solve one can be used for solving the other. In this context, Theorem~\ref{thm:ratio-equivalence}
shows that in case we learn a probabilistic classifier $s(\mathbf{x})$
which is imperfect up to a monotonic transformation of $r(\mathbf{x})$, then
one can still resort to calibration (i.e., modelling $p_Y(y=s(\mathbf{x}))$)
to compute $r(\mathbf{x})$ exactly.

@@

@@

\bigskip

@@

\section{Applications in HEP}

@@

\begin{figure}[htbp]
\begin{center}
 \includegraphics[height=1.7in]{figures/example-TMVA-BDT.pdf}
 \includegraphics[height=1.7in]{figures/example-TMVA-ROC.pdf}
\caption{Left: an example of the distributions $p_b(\hat s|\nu)$ and $p_s(\hat s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT). Right: the corresponding ROC curve (right) for this and other classifiers. (Figures taken from TMVA manual.)}
\label{fig:tmva}
\end{center}
\end{figure}
%\bigskip

@@

\bigskip

\section{Related work}\label{S:Related}

@@

\cite{Neal:2007zz} also considered the problem of approximating the likelihood function when only a generative model is available. That work sketches a scheme in which one uses a classifier with both $x$ and $\theta$ as an input to serve as a dimensionality reduction map.
The key distinction comes in the handling of $\theta$.  Neal says  ``we cannot use the classifier on real data, since we don't know the correct value for [$\theta]$'' and goes on to outline an approach where one uses regression on a per-event basis to estimate $\hat{\theta}(x)$ and perform the composition $s(x; \hat{\theta}(x))$ (much like profiling).\footnote{Neal considers a lower-dimensional `bottlenecks' $\theta^*=g(\theta)$, which are not essential to the discussion here.}  This can lead to a significant loss of information since (at least in most particle physics examples) a single event carries little information about the true value of $\theta$, though the full data set $D$ may be informative -- for instance, a single observation would not be sufficient to estimate the variance of a distribution, though repeated observations would.  The work of Neal correctly identifies this as an approximation of the target likelihood even in the case of a ideal classifier. In contrast, the approach described here does not eliminate the dependence of the classifier on $\theta$.%
\footnote{As a technical point, in Neal's work, the focus is on approximating the likelihood function (up to a multiplicative constant), which is equivalent to evaluating the ratio  with respect to a fixed $\theta_1$ as in Eq.~\ref{eq:mle_withs}. In Neal's case, the dependence on $\theta$ is eliminated via $s(x; \hat{\theta}(x))$ and the map is constant; however, in this approach the map ratio is explicitly parametrized in terms of $\theta$, so the ratio is important for canceling the corresponding Jacobian factors.} Instead, we embed a parametrized classifier into the likelihood and postpone the evaluation of the classifier to the point of evaluation of the likelihood when $\theta$ is explicitly being tested. This avoids the loss of information that occurs from the regression step $\hat{\theta}(x)$ proposed by Neal and leads to Theorem~1, which is an exact result in the case of an ideal classifier. In both cases, the quality of the classifier is factorized from the calibration of its density, which allows for valid inference even if there is a loss of power due to a non ideal classifier.

>>>>

% Introduction =================================================================

% - Introduction section, mostly the same as before:
%    * likelihood free setup
%    * searches for particles at LHC as an example
%    * brief and clear description of the contribution of this paper
%      (i.e., density ratio estimation with calibrated classifiers,
%             as an alternative to other methods for parameter inference)
% - outline

@@

% Likelihood ratio tests =======================================================

% - Introduce the mathematical problem with notations (was 1.1 and 1.2 before)
% - I would keep some of the comments of 1.3, but I feel some are not necessary

@@

% Approximating likelihood ratios with classifiers =============================

% - Core of the paper (was section 2 + my note on slack)
% - Derive theorem for the ratio of transformed densitites
% - Show that consistent regressors minimizing the squared error loss fulfill conditions of the theorem
% - Discuss that training on signal vs background instead of signal+background vs background is OK because this learns a function which is 1:1. => focus the capacity of the classifier; Generalization to decomposed ratios
% - Discuss about the need of calibration in case of imperfect classifiers (was first paragraphs of section 5)

@@

\begin{proposition} \label{thm:best-classifier}
Let $\mathbf{X} = (X_1, ..., X_p)$ and $Y$ be random input and output variables
with values in ${\cal X} \subseteq \mathbb{R}^p$
and ${\cal Y} = \{0, 1\}$ and mixed joint probability density  function
$p_{\mathbf{X},Y}(\mathbf{x}, y)$. For the squared error loss, the best
regression function $s : {\cal X} \mapsto [0, 1]$, or equivalently the best
probabilistic classifier, is
\begin{equation}
s^*(\mathbf{x}) = \frac{P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{P(Y=0) p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
\end{equation}
\end{proposition}

@@

Equivalently, since $m^{-1}(s^*(\mathbf{x})) = \frac{1 -
s^*(\mathbf{x})}{s^*(\mathbf{x})} = r(\mathbf{x})$,
Proposition~\ref{thm:best-classifier} shows that the likelihood ratio estimation
problem is directly related to probabilistic classification, and that algorithms
proposed to solve one can be used for solving the other. In this context,
Theorem~\ref{thm:ratio-equivalence} shows that in case we learn a probabilistic
classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modelling
$p_Y(y=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.

@@

% Generalized likelihood ratio tests ===========================================

% - mathematical problem (was section 4)
% - maximum likelihood estimator inference from ratios
% - discussion on parameterized classifier and practicalities regarding embedding (was sections 5.2 and 5.3)

@@

@@

% Applications =================================================================

% - typical usage in HEP with a concrete experimental example.
%   * illustrate the method for several classifiers (new)
%    * illustrate the method for several calibration routines (new)
%    * compare with other methods (new)
% - ratio of mixtures and decomposition (special case of generalized likelihood rario tests when parameters only affect the mixture coefficients => no need for parametrized classifiers)
% - measure of particles properties, with a concrete example. Comparison with other likelihood-free inference methods (new)

\section{Applications}

@@

\begin{figure}[htbp]
\begin{center}
 \includegraphics[height=1.7in]{figures/example-TMVA-BDT.pdf}
 \includegraphics[height=1.7in]{figures/example-TMVA-ROC.pdf}
\caption{Left: an example of the distributions $p_b(\hat s|\nu)$ and $p_s(\hat s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT). Right: the corresponding ROC curve (right) for this and other classifiers. (Figures taken from TMVA manual.)}
\label{fig:tmva}
\end{center}
\end{figure}

@@

% Related works ================================================================

\section{Related works}\label{S:Related}

@@

\cite{Neal:2007zz} also considered the problem of approximating the likelihood
function when only a generative model is available. That work sketches a scheme
in which one uses a classifier with both $x$ and $\theta$ as an input to serve
as a dimensionality reduction map. The key distinction comes in the handling of
$\theta$.  Neal says  ``we cannot use the classifier on real data, since we
don't know the correct value for [$\theta]$'' and goes on to outline an approach
where one uses regression on a per-event basis to estimate $\hat{\theta}(x)$ and
perform the composition $s(x; \hat{\theta}(x))$ (much like
profiling).\footnote{Neal considers a lower-dimensional `bottlenecks'
$\theta^*=g(\theta)$, which are not essential to the discussion here.}  This can
lead to a significant loss of information since (at least in most particle
physics examples) a single event carries little information about the true value
of $\theta$, though the full data set $D$ may be informative -- for instance, a
single observation would not be sufficient to estimate the variance of a
distribution, though repeated observations would.  The work of Neal correctly
identifies this as an approximation of the target likelihood even in the case of
a ideal classifier. In contrast, the approach described here does not eliminate
the dependence of the classifier on $\theta$.\footnote{As a technical point,
in Neal's work, the focus is on approximating the likelihood function (up to a
multiplicative constant), which is equivalent to evaluating the ratio  with
respect to a fixed $\theta_1$ as in Eq.~\ref{eq:mle_withs}. In Neal's case, the
dependence on $\theta$ is eliminated via $s(x; \hat{\theta}(x))$ and the map is
constant; however, in this approach the map ratio is explicitly parametrized in
terms of $\theta$, so the ratio is important for canceling the corresponding
Jacobian factors.} Instead, we embed a parametrized classifier into the
likelihood and postpone the evaluation of the classifier to the point of
evaluation of the likelihood when $\theta$ is explicitly being tested. This
avoids the loss of information that occurs from the regression step
$\hat{\theta}(x)$ proposed by Neal and leads to Theorem~1, which is an exact
result in the case of an ideal classifier. In both cases, the quality of the
classifier is factorized from the calibration of its density, which allows for
valid inference even if there is a loss of power due to a non ideal classifier.

% Conclusions ==================================================================
